---
title: "Cat Anatomy Report"
author: "Hank Magan"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output: html_document
abstract: "This document provides a report on data related to cat anatomy, specifically heart weight in grams, body weight in kilograms, and gender"
---

## Model Setup
##### In this section, we established the basic parameters of the model, including cleaning up old data, setting up the working directory, and loading a separate file of useful functions. 

```{r model setup}
rm(list=ls())
setwd("C:/Users/hank/Documents/GitHub/IE3620/Week 1")
source("../myfunctions.R")
```

## Loading Libraries
##### For this study, we are using the MASS package (Modern Applied Statistics with S, from Venables and Ripley). In this package, there are a number of datasets ready for analysis, including the cats dataset.

```{r MASS library}
# load data from the MASS library
library(MASS)
```
```{r MASS warning}
## Warning: package 'MASS' was built under R version 3.5.2
```

## Data Demographics
##### In this section, we look at the basic structure of the data: the head of the data file, which shows the names of the columns and a representative sample of the first several rows of data; the dimension of the data (number of rows and columns); and a descriptive statistical summary of the dataset.

```{r cats import}
head(cats)
```
```{r cats dim}
dim(cats)
```
```{r cats summary}
summary(cats)
```

### Nicer looking summary table
##### We can use tools such as knitr/kable to display data tables in a nicer looking format. In this section, we are displaying the summary descriptive statistics table using these tools. 

```{r knitr table}
knitr::kable(
  summary(cats),
  caption = "Table 1:  Summary table of cat anatomy data"
)
```

## Data munging: rename columns, data in columns
##### “Data munging” is a computer science term that basically means cleaning up the data, usually in fairly destructive ways (not always a bad thing). In this section, we are changing the existing column names (sex, bwt,hwt) to something more readable: Gender, BodyWeight and HeartWeight. Having longer column names has its problems(more typing), but does make your dataset more human-friendly! Likewise, we are changing gender from F/M to Female/Male.

```{r column renaming}
colnames(cats) <- c("Gender", "BodyWeight", "HeartWeight")
names(cats)
```
```{r gender renaming}
cats$Gender <- factor(cats$Gender, levels = c("F", "M"), labels = c("Female", "Male"))
head(cats)
```

## Graphical Representations
##### In exploratory data analysis (EDA), it is useful to create basic plots and graphs to get a sense of the behavior of the data. In this section, we create three types of plots: histograms, pairwise correlation plots, and a simple scatter plot, fitted with a regression line.
### Histograms
##### It is oftentimes useful to see histograms of the numerical data, in this case the two weights. It should be noticed that the data does not follow a normal distribution.
```{r bodyweight histogram}
hist(cats$BodyWeight, main="Histogram of Body weight (kg)")
```
```{r heart weight histogram}
hist(cats$HeartWeight, main="Histogram of Heart weight (g)")
```

##### Pairwise plot of relevant data
##### We now use one of the builtin functions from myfunctions.R to create a pairwise plot. This shows histograms of all of the data, a scatter plot of each pair of data, and a correlation coefficient between each pair.

```{r pairwise plot}
pairs(cats,upper.panel=panel.cor,diag.panel=panel.hist)
```

### Plot of heart weight as a function of body weight (with linear regression line)
##### Finally, we run a scatter plot of heart weight as a function of body weight. Notice that the data is fairly linear, supported by plotting a linear regression line on top of the raw data.

```{r linear regression}
plot(cats$HeartWeight~cats$BodyWeight, main="Simple plot of heart weight \n as a function of body weight", xlab="Body weight (kg)", ylab="Heart weight (g)")
fit<- lm(cats$HeartWeight~cats$BodyWeight)
abline(fit)
```